**Effective application of machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-10-03 submission/manuscript.Rmd
jennas edits

2\. Department of Computer Science and Engineering, University or Michigan, Ann Arbor, MI 49109

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and computational efficiency of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, a decision tree, random forest, and extreme gradient boosting (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 but it was slow to train (83.2 h) and hard to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680, trained much faster (12 min), and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study as the choice will inform expectations of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable between individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of existing data and to apply that knowledge to make predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual with high confidence relies on a model that is built with rigorous methods. However, there are common methodological problems across many of these studies that need to be addressed as the field progresses. These include a lack of transparency in which methods are used and how these methods are implemented; developing and evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize over-optimism for model performance.

Among microbiome researchers, the lack of transparency in justifying a modelling approach has been due to an implicit assumption that more complex models are better because they are more complex. This has resulted in a trend towards using models such as random forest and neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although the more complex models may be better at incorporating non-linear relationships or yield better predictions, they are considered to be black box models because they are not inherently interpretable. These models require post hoc explanations to quantify the importance of each feature in making a prediction and they do not show the structure of how the features are used. Depending on the application of the model, researchers may choose to use different modeling approaches. For example, researchers trying to identify the populations causing a disease would likely want a more interpretable model whereas clinicians may emphasize predictive performance. Although one may feel that they are sacrificing interpretability for performance, that tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. Regardless, it is important for researchers to articulate why they have selected a specific modelling approach or even compare multiple approaches in the same study.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing 7 modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also built four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared their predictive performance, interpretability, and computational efficiency. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a framework that implements a ML pipeline that can be used for any modeling approach, evaluates predictive performance, and demonstrates how to interpret these models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

__Model selection and pipeline construction__ We established a common ML pipeline to train and validate each of the seven models that is based on standard methods within the ML community (REFS by Jenna?)[Figure 1].

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset while the test set was composed of the remaining 20% of the data [Figure 1]. To maintain the distribution of controls and cases that was found with the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%). The training data was used to build the models and the test set was used for evaluating predictive performance.

Second, we trained seven different models using the training data [Table 1]. We selected models with different classification algorithms and regularization methods. Regularization is a technique that discourages overfitting by penalizing the model for learning the training data too well. For regularized logistic regression and SVM with linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM model with linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including decision tree, random forest, and XGBoost and we trained an SVM model with a non-linear kernel.

Third, fitting of these models required selecting appropriate hyperparameters. Hyperparameters are the rules that are learned from the training set in a classification algorithm. For example, in the linear models the regularization term (C) is a hyperparameter that indicates the penalty for overfitting. Similar to regularization term C, all hyperparameters are tuned to find the best model. We selected hyperparameters by performing 100 five-fold cross-validation (CV) repeats on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the best hyperparameter values for each model based on its CV predictive performance using the area under the receiver operating characteristic curve (AUROC) metric [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0.50, where the model’s predictions are no different from random chance. To select the best performing hyperparameter, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and Matlab programming languages are inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization coefficients (C) and was susceptible to performing poorly if the wrong regularization coefficient was assigned to the model by default.

Finally, we trained the full training dataset with the selected hyperparameter values and applied the model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that would indicate that the models were overfit. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and would not indicate a problem with overfitting.

To evaluate the risk for over-optimism of each model, we calculated the range of AUROC values for each model using 100 data-splits. The range among the testing AUROC values within each model varied by `r paste(format(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3), nsmall=3))` on average across the seven models. If we had only done a single split, then there is a risk that we could gotten lucky or unlucky with the performance of the model. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`. These results showed that depending on the data-split, the testing AUROC values showed great variability [Figure 2]. Therefore, it is important to employ the 100 hierarchical data splits that were included in our pipeline to minimize the risk of over-optimism.

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The range among the cross-validation AUROC values within both models at lower sample sizes were much larger than when the full collection of samples was used to train and validate the models. These results showed that because the microbiome data had many features (6920 OTUs), it was important to train the models using appropriate sample sizes to avoid problems with generalizability and over-optimism. Furthermore, it was encouraging that even for a small number of samples, the interquartile range included the median AUROC values for the larger subsetted datasets.

__Interpretation of each ML model.__ Interpretability is the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017]. Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model where we randomly permuted non-correlated features individually and groups of perfectly correlated features across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had a minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`. Effectively, the complexity of the communities was more fully represented in the better performing models [@sze_leveraging_2018; @wirbel_meta-analysis_2019]. At least in this case, it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of tuned hyperparameter settings. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5]. When we subsetted the size of the training dataset, we observed a linear relationship between the size of the dataset and the training times for L2-regularized logistic regression and random forest models [Figure S5].

@@

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations. Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process. In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole to classify them into different categories such as coming from a patient who is healthy or has SRNs. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification. However, the application of ML in microbiome studies is still in its infancy and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.

To address these needs, we developed a framework to train rigorous, transparent, and reproducible models. We benchmarked seven ML models and showed that we can create models that are inherently interpretable and easily trained without losing predictive performance. In terms of predictive performance, the random forest model had the best testing AUROC values compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While random forest took 83.2 hours to train, L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a complex ML model and could only be explained using post-hoc methods such as permutation importance. On the other hand, L2-regularized logistic regression was easier to interpret by ranking the OTUs based on their feature weights and reviewing the signs of these weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the similar AUROC values from the cross-validation and testing.

>>>>

2\. Department of Electrical Engineering and Computer Science, University or Michigan, Ann Arbor, MI 48109

@@

Machine learning (ML) modeling of the human microbiome has the potential to identify microbial biomarkers and aid in diagnosis of many diseases such as inflammatory bowel disease, diabetes, and colorectal cancer. Progress has been made towards developing ML models that predict health outcomes using bacterial abundances, but inconsistent adoption of training and evaluation methods call the validity of these models into question. Furthermore, there appears to be a preference by many researchers to favor increased model complexity over interpretability. To overcome these challenges, we trained seven models that used fecal 16S rRNA sequence data to predict the presence of colonic screen relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We developed a generalizable pipeline to train, validate and interpret the models. To show the effect of model selection, we assessed the predictive performance, interpretability, and training time of L2-regularized logistic regression, L1 and L2-regularized support vector machines (SVM) with linear and radial basis function kernels, decision trees, random forest, and extreme gradient boosted trees (XGBoost). The random forest model performed best at detecting SRNs with an AUROC of 0.695 [IQR 0.651-0.739]  but was slow to train (83.2 h) and difficult to interpret. Despite its simplicity, L2-regularized logistic regression followed random forest in predictive performance with an AUROC of 0.680 [IQR 0.625-0.735], trained faster (12 min), and was more interpretable. Our analysis showed that ML models should be chosen based on the goal of the study, as the choice will inform expectations of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow, there is an increasing desire to use microbiome data to diagnose diseases. However, the structure of the human microbiome is remarkably variable AMONG individuals to the point where it is often difficult to identify the bacterial populations that are associated with diseases using traditional statistical models. This variation is likely due to the ability of many bacterial populations to fill the same niche such that different populations cause the same disease in different individuals. Furthermore, a growing number of studies have shown that it is rare for a single bacterial species to be associated with a disease. Instead, subsets of the microbiome account for differences in health. Traditional statistical approaches do not adequately account for the variation in the human microbiome and typically consider the protective or risk effects of each bacterial population individually. Recently, machine learning models (ML) have grown in popularity among microbiome researchers because of the large amount of data that can now be generated and because the models are effective at accounting for the interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @ai_systematic_2017; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without a separate held-out test data; large variation between the predictive performance on different folds of cross-validation; and large variation between cross-validation and testing performances. Nevertheless, the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets  [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019]. More work is needed to further improve reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a modelling approach has been due to an implicit assumption that more complex models are better. This has resulted in a trend towards using models such as random forest and deep neural networks [@galkin_human_2018; @reiman_using_2017; @fioravanti_phylogenetic_2017;  @geman_deep_2018; @zackular_human_2014] over simpler models such as logistic regression or other linear models [@mossotto_classification_2017; @wirbel_meta-analysis_2019; @thomas_metagenomic_2019]. Although complex models may in some cases, capture important non-linear relationships and therefore yield better predictions, they may result in black boxes that lack interpretibility. Such models require post hoc explanations to quantify the importance of each feature in making predictions. Depending on the goal of the modeling, researchers may choose to use different approaches. For example, researchers trying to identify the populations of microbiota associated wwith disease may desire a more interpretable model, whereas clinicians may emphasize predictive performance. Nonetheless, it is important to understand the tradeoff and accepting that this tradeoff may be minimal [@rudin_please_2018; @rudin_optimized_2018]. It is important for researchers to justify their choice of modelling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest and XGBoost. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline, we used data from a previously published study that sought to classifiy individuals as having normal colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome  [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset we developed a ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets.

@@

__Model selection and pipeline construction__ We established a reusable ML pipeline for model selection and evaluation, focusing on seven different commonly used supervised learning algorithms [Figure 1].

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had normal colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals of which 52 had an SRN (54%). The training data was used to build and select the models and the test set was used for evaluating the model.

We trained seven different models using the training data [Table 1]. We focused on different classification algorithms and regularization methods. Regularization helps to prevent overfitting by penalizing a model that fits the training data too well. For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM with a linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including a decision tree, a random forest, and XGBoost and an SVM with a non-linear kernel.

Model selection required tuning hyperparameters. Hyperparameters are parameters which need to be specified or tuned by the user, in order to train a model for a specific modeling problem. For example, when using regularization, (C) is a hyperparameter that indicates the penalty for overfitting. Hyperparameters are tuned using the training data to find the best model. We selected hyperparameters by performing repeated five-fold cross-validation (CV) on the training set [Figure 1]. The five-fold CV was also stratified to maintain the overall case and control distribution. We chose the hyperparameter values that led to the best average CV predictive performance using the area under the receiver operating characteristic curve (AUROC) [Figure S1 and S2]. The AUROC ranges from 1.0, where the model perfectly distinguishes between cases and controls, to 0, where the model’s predictions are perfectly incorrect. An AUROC value of 0.5 indicates that model's predictions are no differect than random. To select hyperparameters, we performed a full grid search for hyperparameter settings when training our models. Default hyperparameter settings in previously developed ML packages in R, Python, and MATLAB programming languages may be inadequate for effective application of classification algorithms and need to be optimized for each new ML task. In the example of L1-regularized SVM with linear kernel [Figure S1], the model showed large variability between different regularization strengths (C).

Once hyperparameters were selected, we trained the model using the full training dataset and applied the final model to the held-out data to evaluate the testing predictive performance of each model. The data-split, hyperparameter selection, training and testing steps were repeated 100 times to obtain a reliable and robust interpretation of model performance [Figure 1].

__Predictive performance and generalizability of the seven models.__  We evaluated the predictive performance of the seven models to classify individuals as having normal colons or SRNs [Figure 2]. The random forest model had significantly higher test AUROC values than the other models for detecting SRNs (Wilcoxon rank sum test, p < 0.01). The median AUROC of the random forest model was `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Random_Forest"))$test_aucs), 3))`). L2-regularized logistic regression, XGBoost, L2-regularized SVM with linear and radial basis function kernel AUROC values were not significantly different from one another and had median AUROC values of `r paste(format(round(median((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Logistic_Regression"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="XGBoost"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="XGBoost"))$test_aucs), 3))`), `r paste(round(median((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="L2_Linear_SVM"))$test_aucs), 3))`) and  `r paste(round(median((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="RBF_SVM"))$test_aucs), 3))`), respectively. L1-regularized SVM with linear kernel and decision tree had significantly lower AUROC values than the other ML models with median AUROC of `r paste(format(round(median((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3), nsmall=3))` (IQR `r paste(round(IQR((all %>% filter(model=="L1_Linear_SVM"))$test_aucs), 3))`) and `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` (IQR `r paste(round(IQR((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))`), respectively [Figure 2]. Interestingly, these results demonstrate that the most complex model (XGBoost) did not have the best performance and that the most interpretable models (L2-regularized logistic regression and SVM with linear kernel) performed nearly as well non-linear models.

To evaluate the generalizability of each model, we compared the median cross-validation AUROC to the median testing AUROC. If the difference between the cross-validation and testing AUROCs was large, then that could indicate that the models were overfit to the training data. The difference in median AUROCs was `r paste(round(abs(difference_model[difference_index[7],1]), 3))` in L1-regularized SVM with linear kernel, followed by SVM with radial basis function kernel and decision tree with a difference of `r paste(round(abs(difference_model[difference_index[6],1]), 3))` and `r paste(round(abs(difference_model[difference_index[5],1]), 3))`, respectively [Figure 2]. These differences are relatively small and gives us confidence in our estimate of the generalization performance of the models.

To evaluate the variation in the estimated performance, we calculated the range of AUROC values for each model using 100 data-splits. The range among the testing AUROC values within each model varied by `r paste(format(round(mean((all %>% group_by(model) %>% summarise(difference=max(test_aucs)-min(test_aucs)))$difference), 3), nsmall=3))` on average across the seven models. If we had only done a single split, then there is a risk that we could have gotten lucky or unlucky in estimating model performance. For instance, the lowest AUROC value of the random forest model was `r paste(round(min((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` whereas the highest was `r paste(format(round(max((all %>% filter(model=="Random_Forest"))$test_aucs), 3), nsmall=3))`. These results showed that depending on the data-split, the testing performance can vary [Figure 2]. Therefore, it is important to employ multiple data splits when estimating generalization performance.

To show the effect of sample size on model generalizability, we compared cross-validation AUROC values of L2-regularized logistic regression and random forest models when we subsetted our original study design with 490 subjects to 15, 30, 60, 120, and 245 subjects [Figure S3]. The variation in cross-validation performance within both models at lower sample sizes was larger than when the full collection of samples was used to train and validate the models. Because of the high dimensionality of the microbiome data (6920 OTUs), large sample sizes can lead to better models.

__Interpretation of each ML model.__ Interpretability is related to the degree to which humans can understand the reasons behind a model prediction [@miller_explanation_2017; @ribeiro_why_2016; @nori_interpretml:_2019]. Because we often use ML models not just to predict a health outcome but also to identify the biomarkers for a disease, model interpretation becomes crucial for microbiome studies. ML models decrease in interpretability as they increase in complexity. In this study we used two methods to help interpret our models.

@@

Second, to analyze non-linear models we interpreted the feature importance using permutation importance. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a posthoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e, testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTU 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models, shared four of the most important OTUs (OTU 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`.  In this case, similar to previous studies [@sze_leveraging_2018; @wirbel_meta-analysis_2019], it was not possible to distinguish between health and disease using a single OTU. Although permutation analysis allowed us to gauge the importance of an OTU, the analysis was post-hoc (i.e. done using the test data) and these results did not allow us to directly interrogate the models to know whether an OTU was associated with classifying a subject as being healthy or having an SRN.

@@

__The computational efficiency of each ML model.__ We compared the training times of the seven ML models. As expected, the training times increased with the complexity of the model and the number of potential hyperparameter combinations. Also, the linear models trained faster than non-linear models [Figures S1-S2; Figure 5].

@@

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism but are the product of multiple bacterial populations. Traditional statistical approaches are useful for identifying those cases where a single organism is associated with a process. In contrast, ML methods offer the ability to incorporate the structure of the microbial communities as a whole and identify associations between community structure and disease state. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those microbial populations within the communities that are responsible for the classification. However, the application of ML in microbiome studies is still in its infancy and the field needs to develop a better understanding of different ML methods, their strengths and weaknesses, and how to implement them.

To address these needs, we developed an open-sourced framework to ML models. using this pipeline, we benchmarked seven ML models and showed that the tradeoff between model complexity and performance may be less severe than originally hypothesized. In terms of predictive performance, the random forest model had the best AUROC compared to the other six models. However, the second-best model was L2-regularized logistic regression with a median AUROC difference of only 0.015 compared to random forest. While our implementation of random forest took 83.2 hours to train, our L2-regularized logistic regression trained in 12 minutes. In terms of interpretability, random forest is a non-linear ML model and is most often explained using post-hoc methods such as permutation importance. On the other hand, using L2-regularized logistic regression we ranked the importance of each OTU based on their feature weights. Comparing many different models showed us that the most complex model was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that can be generalized to any modeling method that predicts a binary health outcome. We performed a random data-split to create a training set (80% of the data) and a held-out test set (20% of the data), which we used to evaluate predictive performance. We repeated this data-split 100 times to measure the possible variation in predictive performance. During the training, we tuned the model hyperparameters with a repeated five-fold cross-validation. Despite the high number of features microbiome datasets typically have, the models we built with this pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures which create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we took collinearity into consideration by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among correlated features. It is important to know the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. To identify the true underlying microbial factors of a disease, it is crucial to do correlation analyses and further experimentation for biological validation.
